---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

```{r setoptions, echo = T, results = 'hide'}
library(knitr)
opts_chunk$set(messages = F, warning = F)
```

## Loading and preprocessing the data

Load some base packages we need for the analysis, and load the dataset as a table.
Notice that we keep the data as is, i.e., we don't perform any postprocessing at this point.

```{r, results = "hide"}
library(plotly)
library(Hmisc)
library(dplyr)
library(chron)
data <- read.csv("activity.csv")
```

## What is mean total number of steps taken per day?

Group the data by date, summarize the steps within each group,
and plot a histogram of the frequency of these step totals.

```{r}
steps <- data %>% group_by(date) %>% summarize(steps_total = sum(steps))
steps.mean <- mean(steps$steps_total, na.rm = T)
steps.median <- median(steps$steps_total, na.rm = T)

plot <- ggplot(data = steps, aes(steps_total)) +
  geom_histogram() +
  labs(title = "Total number of steps taken each day",
       x = "Total steps",
       y = "Frequency")
ggplotly(plot)
```

On average, `r steps.mean` steps are taken each day, while the median of steps is `r steps.median`.

## What is the average daily activity pattern?

We group the data by interval and compute the mean within each interval group.

```{r}
daily_activity <- data %>% group_by(interval) %>% summarize(average = mean(steps, na.rm = T))
plot <- ggplot(data = daily_activity, aes(interval, average)) +
  geom_line() +
  labs(title = "Average Daily Activity Patern",
       x = "Interval",
       y = "Average Steps")
ggplotly(plot)
```

On average across all the days in the dataset, the `r (daily_activity %>% arrange(-average))$interval[1]`th interval contains the maximum number of steps.

## Imputing missing values

The data contains a non-negligible amount of missing values.
How many there are _exactly_ is what we try to find out now:

```{r, results = "hide"}
data.no_all <- nrow(data)
data.no_na <- nrow(data[is.na(data$steps),])
```

- Total number of rows: `r data.no_all`
- Total number of rows with missing values: `r data.no_na`

So the data we got contains `r signif(data.no_na / data.no_all, 3) * 100`% missing values.
As this is quite a lot, we _impute_ the data with the median of steps taken per day.

```{r}
data.imputed <- data
data.imputed$steps <- impute(data$steps, fun = median)
data.steps_by_date <- tapply(data.imputed$steps, data.imputed$date, sum)

data.steps_by_date.mean <- mean(data.steps_by_date)
data.steps_by_date.median <- median(data.steps_by_date)

plot <- qplot(data.steps_by_date,
              main = "Total number of steps taken each day",
              xlab = 'Total steps',
              ylab = 'Frequency')
ggplotly(plot)
```

Let's compare the mean and median of the original data with those of the imputed data:

```{r}
comparison <- data.frame(Original = c(steps.mean, steps.median),
                         Imputed = c(data.steps_by_date.mean, data.steps_by_date.median),
                         row.names = c("Mean", "Median"))
kable(comparison, caption = "Comparison of original and imputed data")
```

What we observe is that there are numerous periods each day without any collected data.
Therefore the zero steps as the median defaults to zero in these cases.

## Are there differences in activity patterns between weekdays and weekends?

Given the imputed data, let's see how if the walking patterns differ between weekdays and weekends.

```{r}
data.wday <- data.imputed %>% mutate(day_type = as.factor(ifelse(is.weekend(as.Date(date)), "weekend", "weekday")))
data.wday.activity <- data.wday %>%
  group_by(interval, day_type) %>%
  mutate(average = mean(steps))
plot <- ggplot(data = data.wday.activity, aes(interval, average)) +
  geom_line() +
  facet_wrap(~day_type, ncol = 1) + 
  labs(title = "Average Daily Activity Patern",
       x = "Interval",
       y = "Average Steps")
ggplotly(plot)
```
